---
title: "Final Project - COVID 19 DATA"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
library(tidyverse)
library(lubridate)
library(ggrepel)
library(knitr)

# Get URL where Johns Hopkins data stored

url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

# List al files to read in
file_names <- c("time_series_covid19_confirmed_US.csv",
                "time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_US.csv",
                "time_series_covid19_deaths_global.csv")

# Concatenate URL and file names
urls <- str_c(url_in, file_names)

# Read in files and store in variables
# NO GLOBAL DATA USED IN THIS REPORT - DATA READ INTO FILE SO THIS CAN BE EASILY ANALYZED IN FUTURE
us_cases <- read_csv(urls[1])
global_cases <- read_csv(urls[2])
deaths_us <- read_csv(urls[3])
deaths_global <- read_csv(urls[4])

# For all Files move dates from column to rows creating a longer instead of a wide data set
global_cases <- global_cases %>%
  pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long),
                         names_to = "Date",
                         values_to = "Total_Cases")

deaths_global <- deaths_global %>%
  pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long),
                         names_to = "Date",
                         values_to = "Total_Deaths")


pop_global <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv") %>%
  select(Province_State, Country_Region, Lat, Long = Long_, Population)


# Join in additional data, rename columns, ensure dates formated properly
global_data <- global_cases %>%
  full_join(deaths_global) %>%
  rename(Country_Region = `Country/Region`,
         Province_State = `Province/State`) %>%
  left_join(pop_global) %>%
  mutate(Date = mdy(Date))


us_cases <- us_cases %>%
  pivot_longer(cols = -c("UID", "iso2", "iso3", "code3", "FIPS",
                         "Admin2", "Province_State", "Country_Region",
                         "Lat", "Long_", "Combined_Key"),
               names_to = "Date",
               values_to = "Total_Cases")

deaths_us <- deaths_us %>%
  pivot_longer(cols = -c("UID", "iso2", "iso3", "code3", "FIPS",
                         "Admin2", "Province_State", "Country_Region",
                         "Lat", "Long_", "Combined_Key", "Population"),
               names_to = "Date",
               values_to = "Total_Deaths")

us_data <- us_cases %>%
  full_join(deaths_us) %>%
  rename(County = `Admin2`) %>%
  mutate(Date = mdy(Date)) %>%
  select(County:Total_Deaths)

#Clean environment by removing unneeded tables
rm(global_cases, deaths_global, us_cases,
   deaths_us, url_in, file_names, urls, pop_global)

#Create variable for the most recently reported date
last_report_date <- max(us_data$Date)

```

## US Analysis by State


### How Many Covid Cases Does it Take Until a Death is Likely?

&nbsp;

```{r deaths_per_case_state, fig.height = 9, fig.width = 10}

# Get 20 most populated states
most_populous <- us_data %>%
  filter(Date == last_report_date) %>%
  group_by(Province_State) %>%
  summarise(Population = sum(Population, na.rm = TRUE)) %>%
  arrange(desc(Population)) %>%
  head(20)

# Get Covid deaths and case counts based on most recent date
state_data <- us_data %>%
  filter(Date == last_report_date) %>%
  group_by(Province_State) %>%
  summarise(Total_Cases = sum(Total_Cases, na.rm = TRUE),
            Total_Deaths = sum(Total_Deaths, na.rm = TRUE)) %>%
  filter(Total_Deaths > 0) %>%
  mutate(Cases_Per_Death = Total_Cases/Total_Deaths,
         Province_State = fct_reorder(Province_State, Cases_Per_Death)) %>%
  arrange(desc(Cases_Per_Death))

# Plot 20 most populated states from best to worst in terms of Deaths per Case
Best_States_Plot <- state_data %>%
  filter(Province_State %in% most_populous$Province_State) %>%
  ggplot(aes(x = Province_State, y = Cases_Per_Death)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  coord_flip() +
  theme_classic() +
  labs(x = NULL, y = "Number of Cases Per Covid Death", title = "How Many Covid Cases per Covid Death by State", subtitle = "20 Most Populous US States", caption = paste("Death and Covid Counts as of ", last_report_date))

Best_States_Plot


```

&nbsp;
&nbsp;
&nbsp;

```{r state_table_analysis}
kable(state_data)

```

&nbsp;
&nbsp;


## Covid Cases and Deaths by Most Populated Counties



```{r populated_counties}
#Get county level data based on most recent reported date
#Only focus on 50 most populated US counties

most_populous_combined <- us_data %>%
  filter(Date == last_report_date) %>%
  group_by(Combined_Key) %>%
  summarise(Population = sum(Population, na.rm = TRUE)) %>%
  arrange(desc(Population)) %>%
  mutate(Pop_Rank = row_number()) %>%
  filter(Pop_Rank <= 50)

#Filter for proper counties, create cases and deaths per thousand people metric
#Rank Counties by highest case volume
county_data <- us_data %>%
  filter(Date == last_report_date,
         Combined_Key %in% most_populous_combined$Combined_Key) %>%
  group_by(Combined_Key, Province_State) %>%
  summarise(Total_Cases = sum(Total_Cases, na.rm = TRUE),
            Total_Deaths = sum(Total_Deaths, na.rm = TRUE)) %>%
  left_join(most_populous_combined) %>%
  mutate(Cases_Per_Thou = Total_Cases*1000/Population,
         Deaths_Per_Thou = Total_Deaths*1000/Population,
         Thousand_Cases = round(Total_Cases/1000, 2)) %>%
  ungroup() %>%
  arrange(desc(Thousand_Cases)) %>%
  mutate(Cases_Rank = row_number())

```

### Modeling the Number of Deaths (Per Thousand) as a Function of Cases (Per Thousand)

```{r counties_model}
# Model number of covid deaths as a function of covid cases per county
# summarize and print model
# Add in predictions from model to the county data frame

county_model <- lm(Deaths_Per_Thou ~ Cases_Per_Thou, data = county_data)
summary(county_model)

county_data$Prediction <- predict(county_model)
```

&nbsp;


### Which Counties had more Deaths Per Covid Case?


```{r counties_plot, fig.width = 12, fig.height = 13}
# Create a scatter plot showing death and case rates
# label counties with high case volumes
# Add line displaying expected deaths based on case counts

county_plot <- county_data %>%
  ggplot(aes(x = Cases_Per_Thou, y = Deaths_Per_Thou,  color = Province_State)) +
  geom_point(aes(size = Thousand_Cases)) +
  geom_line(aes(y = Prediction), color = "red", alpha = 0.5) +
  geom_text_repel(data = subset(county_data, Cases_Rank <= 20), aes(label = Combined_Key)) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(x = "Number of Covid Cases Per Thousand", y = "Number of Covid Deaths Per Thousand",
       title = "Covid Cases vs Covid Deaths by County", subtitle = "50 Most Populous US Counties",
       caption = paste("Death and Covid Counts as of ", last_report_date),
       color = NULL, size = "Total Cases (k)") +
  guides(fill=guide_legend(nrow=2))

county_plot

```



&nbsp;
&nbsp;

## Final Analysis

### Digging Into County Data

```{r best_and_worst_table}
# Create quick tables showing cases with lowest and highest deaths per cases

best_and_worst <- county_data %>%
  mutate(`Distance From Prediction` = Deaths_Per_Thou - Prediction) %>%
  arrange(`Distance From Prediction`) %>%
  select(County = Combined_Key, `Cases Per Thousand` = Cases_Per_Thou,
         `Deaths Per Thousand` = Deaths_Per_Thou, `Predicted Deaths` = Prediction)

best_counties <- best_and_worst %>%
  head(5)

worst_counties <- best_and_worst %>%
  tail(5) %>%
  arrange(desc(row_number()))

kable(best_counties, caption = "Counties Performing Best in Terms of Death Per Covid Case")

kable(worst_counties, caption = "Counties Performing Worst in Terms of Death Per Covid Case")
    
```


&nbsp;

```{r final_plot, fig.width = 12, fig.height = 5}
# Analyze same 10 counties as above and group by perfomance (best vs worst)

worst_v_best_plot_data <- us_data %>%
  mutate(Category = ifelse(Combined_Key %in% worst_counties$County, "Worst",
                           ifelse(Combined_Key %in% best_counties$County, "Best", "NULL")),
         Month = floor_date(Date, "month")) %>%
  filter(Category != "NULL", Date == Month) %>%
  group_by(Category, Date) %>%
  summarise(Total_Cases = sum(Total_Cases, na.rm = TRUE))
  
# Plot data to see when case volumes spiked for best/worst performing counties

worst_v_best_plot <- worst_v_best_plot_data %>%
  ggplot(aes(x = Date, y = Total_Cases, color = Category)) +
  geom_line(stat = "identity") +
  scale_color_manual(values = c("navyblue", "red")) +
  theme_classic() +
  labs(x = "Date", y = "Total Count of Covid Cases", color = "County Performance",
       title = "High vs Low Performing Counties in Terms of Deaths per Covid Case",
       subtitle = "How does the timeline differ for counties with poor vs good deaths/covid case rates?")

worst_v_best_plot

```



#### Difference in Timelines


This final plot shows drastic differences in the volume of Covid 19 cases for counties who had low deaths per case (blue) and high deaths per case (red).

It appears counties who had high number of deaths per case reached higher volumes of cases earlier in the pandemic. This suggests that treatment was not as effective against positive Covid 19 patients early on in the pandemic versus later on. Therefore, it is likely that having a low rate of deaths per Covid case was not so much a measure of success within the county as much as it was at what stage in the pandemic did cases spike within the county.

&nbsp;

### Bias Areas

I think the conclusion that I suggested in the above paragraph was a belief that I suspected prior to doing this analysis. I attempted to combat my bias by questioning my belief head on to see if my suspicions were correct. However, there are likely numerous factors in play other than when cases spiked that attribute to high death per case rate. Perhaps vaccination rates by state or county would affect these rates later on in the pandemic. The age, race and underlying population make-up of a location is likely a big factor on the death per case rate as well. 


One other belief I held prior to this analysis was that Florida would have higher rates of death per case. This is because of my belief that the age of a typical Florida resident skews older than the age of a typical American. However, when looking at the rates of death per case by state and most populous counties this actually turned out to not be the case. Overall, Florida had one of the better deaths per case ratios amongst heavily populated states. This is most evident in the first visual on page 1.


